Effect of Black Hole Spins on Christodoulou Memory in Binary Mergers and its implication on Observing Gravitational Wave Bursts With Memory in Pulsar Timing Array

### 2019-10-09 notes/Paper.tex
updated paper

\stm{note everywhere I have added ``the'', that ``upto'' should be ``up to'', etc}
\begin{abstract}
	The Christodoulou memory, which is a nonlinear memory effect sourced by the oscillatory component of the gravitational wave stress tensor, produces a growing, nonoscillatory change in the gravitational wave strain. This results in the permanent displacement of a pair of freely falling test masses after the gravitational wave has passed. Merger of supermassive black holes is a powerful source of gravitational wave with Chritodoulou  memory effect. This effect can be potentially seen in pulsar timing observations. Initiatives to observe Burst with memory signal have assumed the signal to be a step function of a given amplitude\cite{Pshirkov2010}, which is valid in  situations like supernova explosion. But events like merger of supermassive black holes happens on the time scale of few days depending upon the mass. With future pulsar timing observation, it would be possible to see the change in Chritodoulou memory over two weeks interval of time. We investigate how the effect of Spin changes the timing residuals over a two weeks period of interval. Since Chritodoulou memory is an accumulated effect during the entire history of the source and while memory contribution during the early inspiral is well described by the post-Newtonian (PN) approximation, the contribution at merger must be obtained using numerical relativity simulations. The corrections up to third PN order (i.~e.~3PN) for the gravitational wave memory from pairs of nonspinning black holes on quasicircular orbits has been computed by Favata \cite{Favata2009}. Here we include the contribution from the individual black-hole spins at leading order which appears at 1.5PN \stm{what is leading order here, if we then say the memory is 3PN?} and calculate the memory effect up to 3PN order. We also compute the memory contribution during
	the merger by using the methods presented in \cite{Favata2009} to calculate the memory content using the oscillatory radiation as input, and
	applying that method using publicly available numerical relativity simulations of spinning binaries. We Finally use this model to compute the timing residuals for a memory growth during two weeks interval of time following the method described in \cite{Pshirkov2010}
\end{abstract}

@@

\begin{figure}
 	\includegraphics[width=3.0in]{../plots/MemoryPlot_nonSpining/q7.pdf}
 	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. This figure shows a comparison for different mass ratios. It can be seen that higher mass ration tends to decrease the Chistodoulou memory. The early inspiral part is computed using post-Newtonian expression and the later part during BBH merger is computer using NR data. The red dashed line shows result from PN expression}
 	\label{fig:differentmassratio}
\end{figure}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AlignedSpinSXSdata/0p99.pdf}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AntialignedSpinSXSdata/m0p94.pdf}
	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. The top figure shows a comparison for different values of dimensionless spin parameter $\chi$. These results are for BBH system with equal spin magnitude for individual black hole and spins aligned with orbital angular momentum. It can be seen that higher spin magnitude tend to increase the Christodoulou memory. The bottom figure shows the anti aligned case.  }
	\label{fig:differntchivalues}
\end{figure}
\section{Observing Memory from merger of Super-massive black holes  }
Nonlinear memory from the merger of super massive black hole binary can be potentially observed in pulsar timing measurement. Gravitational waves, when they pass through the space between earth and pulsar effect the time of arrival of the radio pulses at earth. The modulation in the observed frequency is given by
\begin{equation}
\frac{\Delta \nu}{\nu_{0}} = \frac{1}{2c}\int_{0}^{D}d\lambda \left(e^{i}e^{j}\frac{\partial h_{ij}}{\partial t}\right) \Bigg\vert_{path}
\end{equation}

The pulsar timing measurement primarily measures the difference between the actual pulse arrival time and the time predicted from the spin-down model of a pulsar, which is called the timing residuals. The timing residuals $s(t)$, accumulated during a time interval of length due to passing gravitational waves is given by the following expression 
\begin{equation}
	s(t) = \frac{1}{2}(1+\mu)\left[\left(\int_{0}^{t}d\tau \, h_{+}(\tau)\right)\cos 2\phi + \left(\int_{0}^{t}d\tau \, h_{\times}(\tau)\right)\sin 2\phi\right]
\end{equation}
There are two unknown quantities associated with the timing residuals, the pulsar periods and it's spin down rate. We subtract these two quantities from the expression $s(t)$,

The growth in the nonlinear memory signal amplitude from super massive black hole occurs on the time scale of a few days so the effect on timing residuals can be seen if observed over a few weeks period of time. A method for observing Nonlinear memory signal from super massive black hole mergers using pulsar timing has been proposed in ref\cite{Pshirkov2010}. The authors have shown that the present day pulsar timing precision allows one to detect BWM due to  SMBH merger at distance upto 1 Gpc (for case of equal mass $10^{8}$ solar mass). The details regarding the calculation of the timing residuals can be seen in \cite{Pshirkov2010}, where they consider the memory event as a step function which occurs in the middle of an observation time of 10 years. The method described in \cite{Pshirkov2010} is general and can be applied to case where the memory jump is not at step function. In our analysis we explore how the nonlinear memory grows during two weeks interval to time for different ranges of mass and spin parameter values. We also calculate the timing residuals as done in \cite{Pshirkov2010} and explore how it changes as a function of total mass and spin of black holes as shown in Fig[\ref{fig:MemoryMassVsSpin}].
\par 
Future observations with increased precision would be able to look at the memory growth on even shorter time of few weeks. The memory growth on the time scale of two weeks does not look like a step function, instead it's shape depends upon the spins of the merging black holes. In Fig[\ref{fig:resedualGrowth}].we show the memory growth interval in two weeks interval of time with solid lines for different range of spin and total mass of the binary. The dashed line show the complete memory growth in growth in geometrical units. It can be seen the as the total mass of binary becomes larger the two weeks interval of time shrinks in geometrical units. We have chosen this two weeks interval of time such that the change in memory during this time interval is maximum. For the pulsar timing observation we compute the residuals during this two week interval of time following the method described in \cite{Pshirkov2010}. We show a figure similar to
%%\begin{widetext}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksGeoUnits.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow9MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow11MSun.pdf}
	\caption{The top figure shows the memory growth during merger of a binary with time measured in geometric units with total spin magnitude as indicated above. The two weeks interval of time for a given mass as indicated above. The bottom three figures show the the memory growth in days for Mass $10^{8}M_{\odot}$, $10^{9}M_{\odot}$ and $10^{11}M_{\odot}$ respectively.}
	\label{fig:resedualGrowth}
\end{figure} 
%\end{widetext}

>>>>

\stm{note everywhere I have added ``the'', that ``upto'' should be ``up to'', etc}
\begin{abstract}
	The Christodoulou memory, which is a nonlinear memory effect sourced by the oscillatory component of the gravitational wave stress tensor, produces a growing, nonoscillatory change in the gravitational wave strain. This results in the permanent displacement of a pair of freely falling test masses after the gravitational wave has passed. Merger of supermassive black holes is a powerful source of gravitational wave with Chritodoulou  memory effect. This effect can be potentially seen in pulsar timing observations. Initiatives to observe Burst with memory signal have assumed the signal to be a step function of a given amplitude\cite{Pshirkov2010}, which is valid in  situations like supernova explosion. But events like merger of supermassive black holes happens on the time scale of few days depending upon the mass. With future pulsar timing observation, it would be possible to see the change in Chritodoulou memory over two weeks interval of time. We investigate how the effect of Spin changes the timing residuals over a two weeks period of interval. Since Chritodoulou memory is an accumulated effect during the entire history of the source and while memory contribution during the early inspiral is well described by the post-Newtonian (PN) approximation, the contribution at merger must be obtained using numerical relativity simulations. The corrections up to third PN order (i.~e.~3PN) for the gravitational wave memory from pairs of nonspinning black holes on quasicircular orbits has been computed by Favata \cite{Favata2009}. Here we include the contribution from the individual black-hole spins at leading order which appears at 1.5PN \stm{what is leading order here, if we then say the memory is 3PN?} and calculate the memory effect up to 3PN order. We also compute the memory contribution during
	the merger by using the methods presented in \cite{Favata2009} to calculate the memory content using the oscillatory radiation as input, and
	applying that method using publicly available numerical relativity simulations of spinning binaries. We finally use this model to compute the timing residuals for a memory growth during two weeks interval of time following the method described in \cite{Pshirkov2010}. We find that when the burst with memory( BWM) signals are not assumed to a step function and replaced with the actual memory growth functions can significantly chage the expected results. Our results shows that spin contribution plays an important role while computing the residuals.
\end{abstract}

@@

\begin{figure}
 	\includegraphics[width=3.0in]{../plots/MemoryPlot_nonSpining/q7.pdf}
 	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. This figure shows a comparison for different mass ratios. It can be seen that higher mass ration tends to decrease the Chistodoulou memory. The early inspiral part is computed using post-Newtonian expression and the later part during BBH merger is computer using NR data. The red dashed line shows result from PN expression}
 	\label{fig:differentmassratio}
\end{figure}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AlignedSpinSXSdata/0p99.pdf}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AntialignedSpinSXSdata/m0p94.pdf}
	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. The top figure shows a comparison for different values of dimensionless spin parameter $\chi$. These results are for BBH system with equal spin magnitude for individual black hole and spins aligned with orbital angular momentum. It can be seen that higher spin magnitude tend to increase the Christodoulou memory. The bottom figure shows the anti aligned case.  }
	\label{fig:differntchivalues}
\end{figure}
\section{Observing Memory from merger of Super-massive black holes  }
Nonlinear memory from the merger of super massive black hole binary can be potentially observed in pulsar timing measurement. Gravitational waves, when they pass through the space between earth and pulsar effect the time of arrival of the radio pulses at earth. The modulation in the observed frequency is given by
\begin{equation}
\frac{\Delta \nu}{\nu_{0}} = \frac{1}{2c}\int_{0}^{D}d\lambda \left(e^{i}e^{j}\frac{\partial h_{ij}}{\partial t}\right) \Bigg\vert_{path}
\end{equation}
The pulsar timing measurement primarily measures the difference between the actual pulse arrival time and the time predicted from the spin-down model of a pulsar, which is called the timing residuals. The timing residuals $s(t)$, accumulated during a time interval of length due to passing gravitational waves is given by the following expression 
\begin{equation}
	s(t) = \frac{1}{2}(1+\mu)\left[\left(\int_{0}^{t}d\tau \, h_{+}(\tau)\right)\cos 2\phi + \left(\int_{0}^{t}d\tau \, h_{\times}(\tau)\right)\sin 2\phi\right]
\end{equation}
There are two unknown quantities associated with the timing residuals, the pulsar periods and it's spin down rate. We subtract these two quantities from the expression $s(t)$.

The growth in the nonlinear memory signal amplitude from super massive black hole occurs on the time scale of a few days so the effect on timing residuals can be seen if observed over a few weeks period of time. A method for observing Nonlinear memory signal from super massive black hole mergers using pulsar timing has been proposed in ref\cite{Pshirkov2010}. The authors have shown that the present day pulsar timing precision allows one to detect BWM due to  SMBH merger at distance upto 1 Gpc (for case of equal mass $10^{8}$ solar mass). The details regarding the calculation of the timing residuals can be seen in \cite{Pshirkov2010}, where they consider the memory event as a step function which occurs in the middle of an observation time of 10 years. The method described in \cite{Pshirkov2010} is general and can be applied to case where the memory jump is not at step function. In our analysis we explore how the nonlinear memory grows during two weeks interval to time for different ranges of mass and spin parameter values. We also calculate the timing residuals as done in \cite{Pshirkov2010} and explore how it changes as a function of total mass and spin of black holes as shown in Fig[\ref{fig:MemoryMassVsSpin}].
\par 
Future observations with increased precision would be able to look at the memory growth on even shorter time of two weeks. The memory growth on the time scale of two weeks does not look like a step function, instead it's shape depends upon the spins of the merging black holes. In Fig[\ref{fig:MemoryGrowthTwoweeks}].we show the memory growth during two weeks interval of time with markers on complete memory during for different range of spin and total mass of the binary. The solid line show the complete memory growth in growth in geometrical units. It can be seen the as the total mass of binary becomes larger the two weeks interval of time shrinks in geometrical units. We have chosen this two weeks interval of time such that the change in memory during this time interval is maximum. For the pulsar timing observation we compute the residuals during this two week interval of time following the method described in \cite{Pshirkov2010}. It can be seen in Fig[\ref{fig:MemoryGrowthTwoweeks}] that the memory growth during this two weeks interval of time look very for different mass values. For higher masses ($10^{10}M_{\odot}$), this looks like a straight line. For lower masses this looks like a nonlinear growth. In Fig[\ref{fig:resedualG}] we show the residual before(pre-fit) and after  (post-fit) a quadratic subtraction. It can be seen that the residual growth fits better to a pre-fit residuals 
%%\begin{widetext}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksGeoUnits.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow8MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow9MSun.pdf}
	\includegraphics[width=3.0in]{../plots/PlotfromMathematicaData/MemoryGrowth2weeksInDays10pow11MSun.pdf}
	\caption{The top figure shows the memory growth during merger of a binary with time measured in geometric units with total spin magnitude as indicated above. The two weeks interval of time for a given mass as indicated above. The bottom three figures show the the memory growth in days for Mass $10^{8}M_{\odot}$, $10^{9}M_{\odot}$ and $10^{11}M_{\odot}$ respectively.}
	\label{fig:resedualG}
\end{figure} 
%\end{widetext}
